---
title: "Evaluating cluster analysis methods. The case study of Palaeolithic distribution in Galician territory (NW Iberian Peninsula): source code and data"
author: "Mikel Diaz-Rodriguez and Ramon Fabregas-Valcarce"
date: "April 26th, 2021"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE,
                      echo = TRUE)
```

# Introduction

This document contains code to enable reproducibility of the paper "Evaluating cluster analysis methods. The case study of Palaeolithic distribution in Galician territory (NW Iberian Peninsula)".

The steps indicated in the following lines allow to reproduce the analyses carried out.

# Packages used
### Import the packages

```{r echo=TRUE, results='hide'}
# ipak function: install and load multiple R packages.

ipak <- function(pkg){
  new.pkg <- pkg[!(pkg %in% installed.packages()[, "Package"])]
  if (length(new.pkg)) 
    install.packages(new.pkg, dependencies = TRUE)
  sapply(pkg, require, character.only = TRUE)
}
# install all the packages
packages <- c("spatstat", "sp", "rgeos", "maptools", "GISTools", "ggplot2", "fpc", "plyr", "gridExtra", "dbscan", "tidyverse", "cluster", "factoextra", "gridExtra", "readxl", "NbClust", "percopackage", "rgdal", "igraph")
ipak(packages)
```


# Complete Spatial Randomness
### Import the data used in CSR analysis

```{r, message=FALSE, warning=FALSE, paged.print=FALSE}
# set the working directory
setwd("~/shp")

# set the projection 
BNG = "+init=epsg:25829"

# add the study area 
studyarea <- readShapePoly("Galicia.shp", proj4string = CRS(BNG))

# and check it is worked
plot(studyarea)

# add the sites
sites <- readShapePoints("sites.shp", proj4string = CRS(BNG))

# add the random points
random_sites <- readShapePoints("random_sites.shp", proj4string = CRS(BNG))

# plot the studyarea and the sites and check everything looks ok
plot(sites, col="blue", pch=20, add=T)

# remove any sites with the same grid reference
sites <- remove.duplicates(sites)

# select the points inside the studyarea 
sitesSub <- sites[studyarea,]

# check to see that they have been removed
plot(studyarea)
plot(sitesSub, col="red", pch=20, add=T)

# set a window as the studyarea
window <- as.owin(studyarea)
plot(window)

# create a ppp object
sitesSub.ppp <- ppp(x=sitesSub$Easting,y=sitesSub$Northing,window=window)

# plot the sites and the studyarea
plot(sitesSub.ppp,pch=16,cex=0.5, main="Palaeolithic archaeological sites")
```

### Shapiro-Wilk Test

```{r, message=FALSE, warning=FALSE, paged.print=FALSE}
# Shapiro-Wilk Test
shapiro.test(sites$Easting)
```

### K-S Test

```{r, message=FALSE, warning=FALSE, paged.print=FALSE}
# K-S sites vs random sites
ks.test(sites$Easting,random_sites$Easting)
```

### Kernel density estimation methods for cluster analysis selection

```{r echo=TRUE, results='hide'}
# bw.diggle estimation method
a <- bw.diggle(sitesSub.ppp)
a
plot(a, main="a. Likelihood crooss-validation bw.diggle criterion for smoothing bandwith")
pdf(file = "~output/Figure2a.pdf",width = 5,height = 12)

# bw.ppl estimation method
b <- bw.ppl(sitesSub.ppp)
b
plot(b, main="b. Likelihood cross-validation bw.ppl criterion for smoothing bandwith")
pdf(file = "~output/Figure2b.pdf",width = 5,height = 12)

# bw.scott estimation method
c <- bw.scott(sitesSub.ppp)
c
plot(c, main="c. Bandwith selection in multidimensional smoothing with bw.scott method")
pdf(file = "~output/Figure2c.pdf",width = 5,height = 12)

# bw.frac estimation method
d <- bw.frac(sitesSub.ppp)
d
plot(d, main="d. Bandwith selection rule based on the window geometry with bw.frac method")
pdf(file = "~output/Figure2d.pdf",width = 5,height = 12)

# bw.CvL estimation method
e <- bw.CvL(sitesSub.ppp) 
e
plot(e, main="e. Smoothing bandwith selection based on Cambell's formula with bw.CvL method")
pdf(file = "~output/Figure2e.pdf",width = 5,height = 12)
```

### KDE

Kernel density estimation analysis for the best sigmas established in the previous step.

```{r echo=TRUE, results='hide'}
plot(density(sitesSub.ppp, sigma = 413.4299), main="a. Kernel Density (sigma = 0.41 km)")
points(sites$Easting, sites$Northing, pch = 16, cex = 0.2, col = "black")
pdf(file = "~output/Figure3a.pdf",width = 4,height = 4)

plot(density(sitesSub.ppp, sigma = 5314.439), main="b. Kernel Density (sigma = 5.31 km)")
points(sites$Easting, sites$Northing, pch = 16, cex = 0.2, col = "black")
pdf(file = "~output/Figure3b.pdf",width = 4,height = 4)

plot(density(sitesSub.ppp, sigma = 14180.02), main="c. Kernel Density (sigma = 14.18 km)")
points(sites$Easting, sites$Northing, pch = 16, cex = 0.2, col = "black")
pdf(file = "~output/Figure3c.pdf",width = 4,height = 4)

plot(density(sitesSub.ppp, sigma = 17790.83), main="d. Kernel Density (sigma = 17.79 km)")
points(sites$Easting, sites$Northing, pch = 16, cex = 0.2, col = "black")
pdf(file = "~output/Figure3d.pdf",width = 4,height = 4)

plot(density(sitesSub.ppp, sigma = 25282.11), main="e. Kernel Density (sigma = 25.28 km)")
points(sites$Easting, sites$Northing, pch = 16, cex = 0.2, col = "black")
pdf(file = "~output/Figure3e.pdf",width = 4,height = 4)

plot(density(sitesSub.ppp, sigma = 56825.16), main="f. Kernel Density (sigma = 56.83 km)")
points(sites$Easting, sites$Northing, pch = 16, cex = 0.2, col = "black")
pdf(file = "~output/Figure3f.pdf",width = 4,height = 4)
```

### Quadrat Test

```{r echo=TRUE, results='hide'}
# plot the points
plot(sitesSub.ppp,pch=16,cex=0.5, main="g. Quadrat Test results")
# count the points in that fall in a 6 x 6 grid overlaid across the window
plot(quadratcount(sitesSub.ppp, nx = 6, ny = 6),add=T,col="red")
pdf(file = "~output/Figure3g.pdf",width = 8,height = 8)

# if the p-value of Chi-Squared test is > 0.05, then we can reject a null hyphothesis that says "there is no complete spatial randomness in our data". If p-value is > 0.05 then this indicates that we have CSR and there is no pattern in our points. If it is < 0.05, this indicates that we do have clustering in our points.
teststats <- quadrat.test(sitesSub.ppp, nx = 6, ny = 6)

teststats


# the top-left value is the observed count of points; the top-right is the Poisson expected number of points; the bottom value is the Pearson residual value, or (Observed - Expected) / Sqrt(Expected).
plot(sitesSub.ppp,pch=16,cex=0.5, main="h. Quadrat Test Stats results")
plot(teststats, add=T, col = "red")
pdf(file = "~output/Figure3h.pdf",width = 8,height = 8)
```

### K, L and G Functions

```{r echo=TRUE, results='hide'}
# set the simulations
sims <- 99
# to get 95% envelope
nrank <- round((sims + 1) / 100 * 2.5, 0) 

# Kfunction sites
Kfunction <- envelope(sitesSub.ppp,Kest, nsim=sims,rank=nrank, correction="best") 
# Homogeneous
plot(Kfunction,main="a. K Function (homogeneous)",legend=FALSE)
pdf(file = "~output/Figure4a.pdf",width = 5,height = 12)
# Inhomogeneous
KinhomFunction <- envelope(sitesSub.ppp,Kinhom, nsim=sims,rank=nrank, correction="best") 
plot(KinhomFunction,main="b. K Function (inhomogeneous)",legend=FALSE)
pdf(file = "~output/Figure4b.pdf",width = 5,height = 12)

# LFunction sites
LFunction <- envelope(sitesSub.ppp,Lest, nsim=sims,rank=nrank, correction="best")
# Homogeneous
plot(LFunction,main="c. L Function (homogeneous)",legend=FALSE)
pdf(file = "~output/Figure4c.pdf",width = 5,height = 12)
# Inhomogeneous
LinhomFunction <- envelope(sitesSub.ppp,Linhom, nsim=sims,rank=nrank, correction="best")
plot(LinhomFunction,main="d. L Function (inhomogeneous)",legend=FALSE) 
pdf(file = "~output/Figure4d.pdf",width = 5,height = 12)

# GFunction sites
GFunction <- envelope(sitesSub.ppp,Gest, nsim=sims,rank=nrank, correction="best") 
# Homogeneous
plot(GFunction,main="e. G Function (homogeneous)",legend=FALSE) 
pdf(file = "~output/Figure4e.pdf",width = 5,height = 12)
# Inhomogeneous
GinhomFunction <- envelope(sitesSub.ppp,Ginhom, nsim=sims,rank=nrank, correction="best") 
plot(GinhomFunction,main="f. G Function (inhomogeneous)",legend=FALSE) 
pdf(file = "~output/Figure4f.pdf",width = 5,height = 12)
```

### Other statistical tests

Different statistical tests to analyse the sample of sites.

```{r echo=TRUE, results='hide'}
# X2 test sites
chisq.test(sites$Easting,sites$Northing)

# DCLF Test 
dclf.test(sitesSub.ppp)

# Clark and Evans Test
clarkevans.test(sitesSub.ppp)

# Hopkins-Skellam Test 
hopskel.test(sitesSub.ppp, alternative=c("clustered"),method=c("MonteCarlo"),nsim=999)
```

# Kmeans analysis
### Import data

```{r echo=TRUE, results='hide'}
# set the working directory
setwd("~/xls")
data <- read_excel("data.xls")
View(data)

# remove any missing value that might be present in the data
df <- na.omit(data)

# start by scaling/standardizing the data 
df <- scale(df)
head(df) 

# for computing a distance matrix between the rows of a data matrix for Euclidean distance
distance <- get_dist(df) 
```

### Optimal number of clusters

```{r echo=TRUE, results='hide'}
# all indices included to be calculated
resnumclust <- NbClust(df, distance = "euclidean", min.nc=2, max.nc=24, method = "kmeans", index = "alllong")
fviz_nbclust(resnumclust)
```

### Cluster distribution

```{r echo=TRUE, results='hide'}
# clusters
k2 <- kmeans(df, centers = 2, nstart = 25) #test with 2 clusters
k5 <- kmeans(df, centers = 5, nstart = 25) #test with 5 clusters
k7 <- kmeans(df, centers = 7, nstart = 25) #test with 7 clusters
k13 <- kmeans(df, centers = 13, nstart = 25) #test with 13 clusters

# plot the clusters
p1 <- df %>%
  as_tibble() %>%
  mutate(cluster = k2$cluster,
         state = row.names(data)) %>%
  ggplot(aes(data$Easting, data$Northing, color = factor(cluster), label = state)) +
  geom_point()  + theme_bw() + labs(title = "Kmeans (N=2)",y= "UTM Y", x = "UTM X",colour = "Cluster") 

p2 <- df %>%
  as_tibble() %>%
  mutate(cluster = k5$cluster,
         state = row.names(data)) %>%
  ggplot(aes(data$Easting, data$Northing, color = factor(cluster), label = state)) +
  geom_point() + theme_bw() +labs(title = "Kmeans (N=5)",y= "UTM Y", x = "UTM X",colour = "Cluster") 

p3 <- df %>%
  as_tibble() %>%
  mutate(cluster = k7$cluster,
         state = row.names(data)) %>%
  ggplot(aes(data$Easting, data$Northing, color = factor(cluster), label = state)) +
  geom_point() + theme_bw() +labs(title = "Kmeans (N=7)",y= "UTM Y", x = "UTM X",colour = "Cluster") 

p4 <- df %>%
  as_tibble() %>%
  mutate(cluster = k13$cluster,
         state = row.names(data)) %>%
  ggplot(aes(data$Easting, data$Northing, color = factor(cluster), label = state)) +
  geom_point() + theme_bw() +labs(title = "Kmeans (N=13)",y= "UTM Y", x = "UTM X",colour = "Cluster") 

# plot comparing the 4 clusters (N=2/N=13)
grid.arrange(p1, p2, p3, p4, nrow = 2)
pdf(file = "~output/Figure4f.pdf",width = 6,height = 10)
```

# DBSCAN analysis
### Import data

```{r echo=TRUE, results='hide'}
# set the working directory
setwd("~/shp")

# EPSG string to help set the projection 
BNG = "+init=epsg:25829"
# read the data in
studyarea <- readShapePoly("Galicia.shp", proj4string = CRS(BNG))
# check it is worked
plot(studyarea)

# add the sites
sites <- readShapePoints("sites.shp", proj4string = CRS(BNG))
# plot the studyarea and the sites and check everything looks ok
plot(sites, col="blue", pch=20, add=T)

# remove any sites with the same grid reference
sites <- remove.duplicates(sites)
# select the points inside the studyarea 
sitesSub <- sites[studyarea,]
# check to see that they have been removed
plot(studyarea)
plot(sitesSub, col="red", pch=20, add=T) 

# set a window as the studyarea
window <- as.owin(studyarea)
plot(window)

# create a ppp object
sitesSub.ppp <- ppp(x=sitesSub$Easting,y=sitesSub$Northing,window=window)

# this option is to check the coordinate reference system of the spatial polygon:
crs(studyarea)

# extract the points from the spatial points data frame
sitesPoints <- data.frame(x=sitesSub$Easting,y=sitesSub$Northing)
```

### Evaluation of eps

```{r echo=TRUE, results='hide'}
# indicate min_pts
min_pts <- 4


# evaluate the eps and chose the value
A <- dbscan::kNNdist(sitesPoints, k = min_pts)
Eps <- A[order(A)]
plot(Eps)

abline(h=6000, col = "red", lty=2) + title("a.Evaluation of eps")
pdf(file = "~output/Figure7a.pdf",width = 5,height = 12)
```

### Create DBSCAN

```{r echo=TRUE, results='hide'}
# run the dbscan analysis for different eps
db1 <- fpc::dbscan(sitesPoints, eps = 1000, MinPts = 4)

db2 <- fpc::dbscan(sitesPoints, eps = 6000, MinPts = 4) 

db3 <- fpc::dbscan(sitesPoints, eps = 10000, MinPts = 4) 

db4 <- fpc::dbscan(sitesPoints, eps = 20000, MinPts = 4)

# create a ggplot2 object from the data
cluster1 <- as.factor(as.character(db1$cluster))
p1 <- ggplot(data=sitesPoints, aes(x=sitesSub$Easting, y=sitesSub$Northing, colour=cluster1, factor(cluster1))) + 
  geom_point()+ theme_bw() + labs(title = "DBSCAN Cluster (eps = 1000)", y= "UTM Y", x = "UTM X") + 
  scale_color_manual(values = c("oldlace", "#2E9FDF", "#FC4E07","peru", "seagreen", "seashell3", "violetred", "yellow1", "tan2", "slateblue4", "plum2", "mediumspringgreen", "lightsteelblue1", "lightsalmon1", "khaki4", "lightcoral", "mediumspringgreen", "lightyellow1", "maroon3", "olivedrab4", "midnightblue", "navajowhite4")) + labs(colour = "Cluster Number")
plot(p1)

cluster2 <- as.factor(as.character(db2$cluster))
p2 <- ggplot(data=sitesPoints, aes(x=sitesSub$Easting, y=sitesSub$Northing, colour=cluster2)) + 
  geom_point()+ theme_bw() + labs(title = "DBSCAN Cluster (eps = 6000)",y= "UTM Y", x = "UTM X") + 
  scale_color_manual(values = c("oldlace", "#2E9FDF", "#FC4E07","peru", "seagreen", "seashell3", "violetred", "yellow1", "tan2", "slateblue4", "plum2", "mediumspringgreen", "lightsteelblue1", "lightsalmon1")) + labs(colour = "Cluster Number")
plot(p2)

cluster3 <- as.factor(as.character(db3$cluster))
p3 <- ggplot(data=sitesPoints, aes(x=sitesSub$Easting, y=sitesSub$Northing, colour=cluster3)) + 
  geom_point()+ theme_bw() + labs(title = "DBSCAN Cluster (eps = 10000)",y= "UTM Y", x = "UTM X") + 
  scale_color_manual(values = c("oldlace", "#2E9FDF", "#FC4E07","peru", "seagreen", "seashell3", "violetred", "yellow1", "tan2", "slateblue4")) + labs(colour = "Cluster Number")
plot(p3)

cluster4 <- as.factor(as.character(db4$cluster))
p4 <- ggplot(data=sitesPoints, aes(x=sitesSub$Easting, y=sitesSub$Northing, colour=cluster4)) + 
  geom_point() + theme_bw() + labs(title = "DBSCAN Cluster (eps = 20000)",y= "UTM Y", x = "UTM X") + 
  scale_color_manual(values = c("oldlace", "#2E9FDF", "#FC4E07","peru")) + labs(colour = "Cluster Number")
plot(p4)


grid.arrange(p1, p2, p3, p4, nrow = 2)
pdf(file = "~output/Figure7b-e.pdf",width = 8,height = 12)
```

# Percolation analysis
### Import data

```{r echo=TRUE, results='hide'}

# set the working directory. The process create new folders in that directory for the results
setwd("~/percolation")

# import sites
data <- read_excel("data.xls")

# import studyarea
Gal_shape <- readOGR(dsn="Galicia.shp", layer="Galicia")
```

### Analysis

```{r echo=TRUE, results='hide'}
# percolation
percolate (data, ,40, 2, 1, 50, 1000)

# create maps
map_name <- "Percolation distance"
dpi <- 300

# create the cluster maps
mapClusters(Gal_shape,map_name,"cluster",dpi)

# create graphs
plotClustFreq ("analysis_by_radius.csv")
```